{\bf \huge Wikiplia}:\\ The Free Programming Language\\ That Anyone Can Edit

### 2007-03-19 wikiplia/paper/wikiplia.tex
fix typo in license ;)

One of the most cherished social principles of mankind is
freedom,\z\wcite{Freedom} in its many incarnations. More recently,
freedom has become an important principle in computer science as well,
with the introduction of Free Software licenses such as the GNU
GPL,\z\wcite{GNU General Public License} extensible markup languages
such as XML,\z\wcite{XML} the ability expicitly deallocate memory
with the {\tt free(3)} library call, and the widespread availibility of
Free Herbal V1agara on the World Wide Web.\z\wcite{Sildenafil}

@@

As a concrete example, suppose there are two programmers called K and
R. Each is modifying the GCC with the purpose of adding a new
character constant, \verb+'\c'+. K and R start at revision 100 of the
GCC. K finds the case analysis for parsing character constants:
%
\begin{code}
/* REVISION 100 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  default: abort("bad char constant");
\}
\end{code}
%
He adds a case for his extension, without using the extension, and
checks this in as revision 101.
%
\begin{code}
/* REVISION 101 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 257;
  default: abort("bad char constant");
\}
\end{code}
%
Meanwhile, R has similar (but not identical) inspiration and modifies
his copy of the compiler:
%
\begin{code}
/* REVISION 100 (R) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 8675309;
  default: abort("bad char constant");
\}
\end{code}
%
He does not commit his code because he is wary of the time-consuming
conflict resolution phase and is late for a date with K's estranged
wife who is fed up with K's all-night hacking binges. He takes off in
his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler for a
night on the town, believing that a healthy well-rounded programmer
spends more or less equal nights basking in the pale amber glow of the
teletype as waking up with a few missing teeth naked and norovirused
in some midtown alleyway with his wallet barely out of reach but empty
anyway, having amply exercised Freedom~\fbeer.

@@

Based on these scenarios we conclude that extant social measures such
as revision control conventions are not enough to guarantee freedom.
Even if we think these situations are implausible in the hands of
well-intentioned, well-mannered software engineers, we wish for our
software to remain free even when in the hands of nefarious and crafty
factions who would seek to fracture\wcite{Fork (software development)}
our free software community. We therefore need a technological and
legal solution that forces the entire development history to be
available. We call this technology and its license {\em Total
Information Awareness} after the successful project of the US
Government with the same name.\z\wcite{Information Awareness Office}

@@

XML documents are universally parseable.\z\wcite{Parsing} However,
they are difficult to write and read. Therefore, as usual\wcite{RELAX
NG} we create a new syntax by which humans can write and read
documents and which the computer automatically parses and converts to
the easily parsable XML syntax and back to the new syntax, reducing
complexity.\z\footnote{For efficiency, the Wikiplia implementation is
optimized to lazily perform these translations, so that the document
is never represented in XML form.} This compact syntax is based upon
parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
parenthesis is not {\em named} as in XML, which makes parsing
difficult because the computer must {\em guess} which parenthesis
belongs to which other parenthesis. We therefore call this compact
syntax XESP because it basically {\em reads the programmer's
mind}\wcite{Extra-sensory perception} to guess what the name of the
closing parenthesis should be. The grammar for XESP is given in
Figure~\ref{fig:xesp} along with the translation to XML documents.
From now on, we use the XESP syntax in our examples.

@@

@@

II. Version Idenfitication and Invalid Licenses

@@

END INVINCIBLE SECTION 4\footnote{This clause makes the license
``viral'' like the GNU GPL, so that freedom is preserved in all
descendents of the software.}

>>>>

One of the most cherished social principles of mankind is
freedom,\z\wcite{Freedom} in its many incarnations. More recently,
freedom has become an important principle in computer science as well,
with the introduction of Free Software licenses such as the GNU
GPL,\z\wcite{GNU General Public License} extensible markup languages
such as XML,\z\wcite{XML} the ability explicitly deallocate memory
with the {\tt free(3)} library call, and the widespread availability of
Free Herbal V1agara on the World Wide Web.\z\wcite{Sildenafil}

@@

As a concrete example, suppose there are two programmers called K and
R. Each is modifying the GCC with the purpose of adding a new
character constant, \verb+'\c'+. K and R start at revision 100 of the
GCC. K finds the case analysis for parsing character constants:
%
\begin{code}
/* REVISION 100 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  default: abort("bad char constant");
\}
\end{code}
%
He adds a case for his extension, without using the extension, and
checks this in as revision 101.
%
\begin{code}
/* REVISION 101 (K) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 257;
  default: abort("bad char constant");
\}
\end{code}
%
Meanwhile, R has similar (but not identical) inspiration and modifies
his copy of the compiler:
%
\begin{code}
/* REVISION 100 (R) */
switch(ch) \{
  case 'n': return '\\n';
  case 'r': return '\\r';
  \vdots
  case 'c': return 8675309;
  default: abort("bad char constant");
\}
\end{code}
%
He does not commit his code because he is wary of the time-consuming
conflict resolution phase and is late for a date with K's estranged
wife who is fed up with K's all-night hacking binges. He burns rubber
in his 2007 Honda Civic\wcite{Honda Civic} with aftermarket spoiler
for a night on the town, believing that a healthy well-rounded
programmer spends more or less equal nights basking in the pale amber
glow of the teletype as waking up with a few missing teeth naked and
norovirused in some midtown alleyway with his wallet barely out of
reach but empty anyway, having amply exercised Freedom~\fbeer.

@@

Based on these scenarios we conclude that extant social measures such
as revision control conventions are not enough to guarantee freedom.
Even if we think these situations are implausible in the hands of
well-intentioned, well-mannered software engineers, we wish for our
software to remain free even when in the hands of nefarious and crafty
factions who would seek to fracture\wcite{Fork (software development)}
our free software community. We therefore need a technological and
legal solution that forces the entire development history to be
available. Keeping with Freedom~\fusa, we call this technology and its
license {\em Total Information Awareness} after the successful project
of the US Government with the same name.\z\wcite{Information Awareness
Office}

@@

XML documents are universally parseable.\z\wcite{Parsing} However,
they are difficult to write and read. Therefore, as usual\wcite{RELAX
NG} we create a new syntax by which humans can write and read
documents and which the computer automatically parses and converts to
the easily parsable XML syntax and back to the new syntax, reducing
complexity.\z\footnote{For efficiency, the Wikiplia implementation is
optimized to lazily perform these translations, so that the document
is never represented in XML form.} This compact syntax is based upon
parentheses rather than tags: The XML document $\tag{list}{X_1\ X_2}$
is instead written ${\tt(}X_1\ X_2{\tt)}$. Note that the closing
parenthesis is not {\em named} as in XML, which makes parsing
difficult because the computer must {\em guess} which parenthesis
belongs to which other parenthesis. We therefore call this compact
syntax XESP because it basically {\em reads the programmer's
mind}\,\wcite{Extra-sensory perception} to guess what the name of the
closing parenthesis should be. The grammar for XESP is given in
Figure~\ref{fig:xesp} along with the translation to XML documents.
From now on, we use the XESP syntax in our examples.

@@

Wikiplia is implemented as a World Wide Web Home-Site, which allows
for easy access from any location.

The system is implemented as a Standard~ML\wcite{Standard ML} program
of approximately 1,000 lines.\z\footnote{This count does not include
general purpose libraries, such as a networking library.} This program
is designed to be minimal: it consists of a web server, a revision
control system, and facilities for evaluating XESP documents. It also
contains a very minimal bootstrapping ``compiler'' for XESP documents,
with its boots manually strapped. From this tiny core we then develop
the remainder of Wikiplia using Wikiplia itself.

Some may balk at the choice of Standard~ML, as the language is
miserably non-free: First, while many of the major implementations are
GNU or BSD-licensed,\z\wcite{BSD licenses} all are implemented in
Standard~ML itself, yielding the bootstrapping problem described
earlier. Second, the mathematical Definition of Standard ML book is
only available as a copyrighted publication of MIT
Press,\z\wcite{MIT Press} not even {\em ostensibly} in a free
manner. However, the performance considerations of the server and
evaluator---and the lack of suitable free alternatives---force us to
settle for such subjugation.

\subsubsection{Web Server}

The web server's job is simple. It runs in a loop, accepting a single
connection, setting up an initial environment $\G$ for evaluation
(which associates the symbols {\sf request.url}, {\sf request.ip} and
{\sf request.time} with the appropriate values), and then evaluating
the document contained at the head of the key ``main'' in the
repository. The result of that evaluation is sent back to the web
browser as a string.\z\footnote{The HTTP result code may be modified
if, for example, the resulting document is a HTTP redirect to another
URL.} The server knows nothing else about how Wikiplia works.

The web server is single-threaded (so each request must finish
before the next is handled) because we desire global atomicity
(Section~\ref{sec:bootstrap}).

\subsubsection{Revision Control}

The revision control system stores XESP documents and their history.
This is mostly a straightforward implementation of the imperative {\tt
cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
revision control systems, we need to support very large numbers of
revisions with small edits (since every change is saved), so the
implementation is engineered to make storing small revisions very
cheap. Particularly, revisions are aggressively compressed by only
storing the newest revision directly, and then a series of difference
``plans'' that describe how to get the next older revision from the
current one. We compute optimal plans using an efficient minimal edit
distance\wcite{Levenshtein distance} calculation at the
token\wcite{Token (parser)} level. As of revision~\currentrevision,
the database is only 236 kilobytes.\z\wcite{Kilobyte}

\subsubsection{Document Evaluation}

The Wikiplia implementation also has facilities for parsing and
evaluating XESP documents. These are a direct implementation of the
evaluation rules in Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.

\subsubsection{Bootstrapping}

The key ``main'' of Wikiplia is responsible for decomposing the URL
and acting upon it however is appropriate. The goal of the
bootstrapping process is to make Wikiplia
self-sufficient\wcite{Self-sufficiency} in the sense that the language
can be edited from the web site implemented by ``main.'' To do so, the
web site needs to be able to present the user with an edit box
containing the current source of the ``main'' key and the ability to
save his code into the database, overwriting the ``main'' key, in
order to add functionality.

The initial implementation of ``main'' is provides for the simple
ability to edit, save, and view the current version of keys in the
repository. These three actions are encoded as the URLs {\tt
/edit/key}, {\tt /save/key}, and {\tt /view/\_/key}. The view action
is straightforward. The edit action displays an HTML
textarea\wcite{Text box} containing the current value of the key and a
button that submits the changes to the {\tt save} url. The save action
is the most complex. First, the submitted document is saved as the new
version of the key. Then, if the key is of the form {\tt
base.extension}, the database is checked to see if there is a key
called {\tt extension:compile}. If so, the XESP document that is there
is applied to the input document (a string) to produce a document that
is saved at the key {\tt base}. This allows us to develop languages
that are automatically compiled when saved.

\begin{figure}
\begin{center}
\verb+(lambda 's '(parse (xcase s 'no '(h _ h))))+
\end{center}
\caption{The initial bootstrapping compiler.} \label{fig:b:compile.b}
\end{figure}

This initial functionality is implemented directly in the XESP
language, whose extension is {\tt b}; the bootstrapping ``compiler''
is just the built-in parser (Figure~\ref{fig:b:compile.b}).

Arranging that the repository contain the correct keys to make this
work is subtle. A small initialization phase sets up:
\begin{itemize}
\item {\tt b:compile.b}$\quad$ 
      The source code (a string) in Figure~\ref{fig:b:compile.b}
\item {\tt b:compile}$\quad$ 
      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt b:compile.b} yields {\tt b:compile}
\item {\tt main.b}$\quad$ 
      The source code of the original ``main'' program
\item {\tt main}$\quad$
      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt main.b} yields {\tt main}.
\end{itemize}

We do this by {\em anticipating}, during the initialization process,
the {\em meaning} of {\tt b:compile} so that we can perform that
action (parsing) on {\tt b:compile.b} to produce {\tt b:compile}. This
can only be achieved by fiat and this is the essence of bootstrapping.

After this minimal initialization, we can then exclusively use the web
interface to develop and extend Wikiplia.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=0.92\textwidth]{ss_homepage}
\end{center}
\caption{Screenshot of the Wikiplia home page as of revision
\currentrevision.} \label{fig:sshomepage}
\end{figure*}

\begin{figure*}[t]
\begin{center}
% \includegraphics[width=0.92\textwidth]{ss_history}
\end{center}
\caption{Screenshot of the history information for {\tt main-go.w}.
} \label{fig:sshistory}
\end{figure*}

@@

II. Version Identification and Invalid Licenses

@@

END INVINCIBLE SECTION 4\footnote{This clause makes the license
``viral'' like the GNU GPL, so that freedom is preserved in all
descendants of the software.}
